---
title: "Introduction to Biomedical Data Science / Homework-4 (Week-5)"
author: "Aysu Aydın"
date: "February 22, 2024"
output:
    html_document:
        toc: true
        toc_depth: 2
        number_sections: true
---


In this homework you will analyze a large data set that contains information regarding flights departing from New York City airports.


```{r}

## Install it if you don't have it
# install.packages("nycflights13")

```

# Load libraries (packages)

- nycflights13
- tidyverse

```{r}
library(nycflights13)
library(tidyverse)

```


# Familiarize yourself with the data set

- Get help on the data set and understand what the indicate
- Show the dimensions of the data
- Look up the structure of the data to understand data types

```{r}
help(flights)
dim(flights)
str(flights)

```


# Use functions that apply to rows (`filter`, `slice`, `arrange` )

- Create a data frame `df1` which should have all entries with `air_time` larger than 150
- Create a data frame `df2` which have flights departing from JFK headed to MIA
- Create a data frame `df3` which should have distances larger than the median value of distances in the whole data frame (hint you can do calculations and use functions inside `filter()` function)
- Create a data frame `df4` subset with the 5 highest values in `dep_delay`
- Create a data frame `df5` by 1) filtering on flights departing from JFK 2) filtering on arrival delays greater than 10, 3) Arranging based on descending departure delays (`dep_delay`) and print the ordered data frame 


```{r}
df1 <- flights %>% 
  filter(air_time >150)

df2 <- flights %>% 
  filter(origin == "JFK" & dest == "MIA")

df3 <- flights %>% 
  filter(distance > median(distance))

df4 <- flights %>% 
  slice_max(order_by = dep_delay, n = 5)

df5 <- flights %>% 
  filter(origin == "JFK") %>% 
  filter(arr_delay > 10) %>% 
  arrange(desc(dep_delay))
  

```


# Use `select()`, `rename()`, and `relocate()`  functions

- Create a data frame `df6` by selecting the following columns: carrier, origin, dest
- Create a data frame `df7` which should include only the columns that contains `arr` in the name
- Create a data frame `df8` by 1) selecting all columns except `year`, `month`, and `day` 2) renaming `arr_time` column as `arrival_time` 3) reorder columns so that first `tailnum` are shown followed by all the other columns
- Create a data frame `df9` by relocating the columns between 6 and 8 to columns 1-3


```{r}
df6 <- flights %>% 
  select(carrier, origin, dest)

df7 <- flights %>% 
  select(contains("arr"))

df8 <- flights %>% 
  select(!c(year, month, day)) %>% 
  rename(arrival_time = arr_time) %>% 
  select(tailnum, everything())

df9 <- flights %>% 
  relocate(6:8, before = 1:3)
  

```


# Use `mutate()`, `transmute()`, `pull()`, `distinct()` functions

Using `flights` data frame as above,

- Create a data frame `df10` with a new called `constant_character` which should be `"abc"`

- Create a data frame `df11` with a new variable called `random_number` which should have random numbers between 1 and 100 (hint you need to make sure random sampling done as many times as there are rows. There is a useful function for getting the row number from a data frame: `nrow()`)

- Using `transmute()`, create a data frame `d12` that should have a variable arr_delay_times5 and this variable should be the result of 5 * arr_delay column

- Create a data frame `df13` by keeping all the distinct rows based on `origin` and `dest` pairs (hint pass `origin` and `dest` arguments into `distinct()` function to force selecting between unique pairs)

- Using pipes, create `vec14` by 1)filter flights originating from JFK, 2) create a new variable `carrier_tailnum` by pasting `carrier` and `tailnum` (use `sep="_"` to concatenate with underscore), 3) pull this new column as vector 4) sample 10 random items from the vector

```{r}
df10 <- flights %>% 
  mutate(constant_character = "abc")

df11 <- flights %>% 
  mutate(random_number = sample(1:100, replace = T, size = nrow(flights)))

df12 <- flights %>% 
  transmute(arr_delay_times5 = 5 * arr_delay)

df13 <- flights %>% 
  distinct(origin, dest)

vec14 <- flights %>% 
  filter(origin == "JFK") %>% 
  mutate(carrier_tailnum = paste(carrier, tailnum, sep = "_")) %>% 
  pull(carrier_tailnum) %>% 
  sample(10)


```


# Apply functions to data subsets

- Create `df15` by grouping flights based on origin city and calculating average `arr_delay`

- Create `df16` by grouping flights based on `carrier` and count how many flights have been done by that carrier (hint `tally()` or `n()`)

- Create `df17` by grouping flights based on both `origin` and `dest` and create three average variables: `avg_distance`, `avg_airtime`, `avg_arrdelay`.

```{r}
df15 <- flights %>% 
  group_by(origin) %>% 
  summarise(avg_delay = mean(arr_delay, na.rm = T))

df16 <- flights %>% 
  group_by(carrier) %>% 
  tally()

df17 <- flights %>% 
  group_by(origin, dest) %>% 
  summarise(avg_distance = mean(distance, na.rm = T),
            avg_airtime = mean(air_time, na.rm = T),
            avg_arrdelay = mean(arr_delay, na.rm = T))

```
